Fractional Types

### 2013-02-24 frac.tex
global revision

\paragraph*{Data.}
We view {{Pi}} as having two levels:  it has traditional values, given by:
%subcode{bnf} include main
% values, v ::= () | left v | right v | (v, v)
\noindent and these are classified by ordinary types:
%subcode{bnf} include main
% value types, b ::= 1 | b + b | b * b 
\noindent 
Types include the unit type {{1}}, sum types {{b1+b2}}, and
products types {{b1*b2}}.  Values includes {{()}} which is the only value of
type {{1}}, {{left v}} and {{right v}} which inject {{v}} into a sum type,
and {{(v1,v2)}} which builds a value of product type. But these values should
be regarded as largely ancillary: we do not treat them as first-class citizens,
and they only occur when observing the effect of an isomorphism.

\paragraph*{Isomorphisms.} The important terms of {{Pi}} are witnesses to
(value) type isomorphisms.  They are also typed, by the shape of the (value)
type isomorphism they witness {{b <-> b}}.  Specifically, they are witnesses to
the following isomorphisms:
%subcode{bnf} include main
%! columnStyle = r@{\hspace{-0.5pt}}r c l@{\hspace{-0.5pt}}l
%swap+ :&  b1 + b2 & <-> & b2 + b1 &: swap+
%assocl+ :&  b1 + (b2 + b3) & <-> & (b1 + b2) + b3 &: assocr+
%identl* :&  1 * b & <-> & b &: identr*
%swap* :&  b1 * b2 & <-> & b2 * b1 &: swap*
%assocl* :&  b1 * (b2 * b3) & <-> & (b1 * b2) * b3 &: assocr*
%dist :&~ (b1 + b2) * b3 & <-> & (b1 * b3) + (b2 * b3)~ &: factor
\noindent Each line of the above table introduces a pair of dual
constants\footnote{where {{swap*}} and {{swap+}} are self-dual} that witness
the type isomorphism in the middle.  These are the base (non-reducible) terms
of the second, principal level of {{Pi}}. Note how the above has two
readings: first as a set of typing relations for a set of constants. Second,
if these axioms are seen as universally quantified, orientable statements,
they also induce transformations of the (traditional) values. The
(categorical) intuition here is that these axioms have computational content
because they witness isomorphisms rather than merely stating an extensional
equality.

@@

It is important to note that ``values'' and ``isomorphisms'' are completely
separate syntactic categories which do not intermix. The semantics of the
language come when these are made to interact at the ``top level'', a new
syntactic category:
%subcode{bnf} include main
% top level term, l ::= c v
\noindent We refer to this as \emph{application}.

@@

Mathematically, {{Pi}} can be interpreted in the category of finite sets and
bijections: each value type denotes a finite set of a size calculated by
viewing the types as natural numbers and each combinator {{c : b1 <-> b2}}
denotes a bijection between the sets denoted by~{{b1}}
and~{{b2}}. Operationally, the semantics of {{Pi}} is given using two
mutually recursive interpreters: one going forward and one going
backwards. The use of {{sym}} switches control from one evaluator to the
other. We will present the operational semantics in the next section along
with the extension with fractional types and values. For now, we state
without proof that the evaluation of well-typed combinators always terminates
and that {{Pi}} is logically reversible, i.e., that for all combinators 
{{c : b1 <-> b2}} and values {{v1 : b1}} and {{v2 : b2}} we have the forward
evaluation of {{c v1}} produces {{v2}} iff the backwards evaluation of 
{{c v2}} produces {{v1}}.

@@

The language presented above, at the type level, models a commutative ringoid
where the multiplicative structure forms a commutative monoid, but the
additive structure is just a commutative semigroup.  Note that the version of
{{Pi}} that include the empty type with its usual laws exactly captures, at
the type level, the notion of a \emph{semiring} (occasionally called a
\emph{rig}) where we replace equality by isomorphism.  This can also be also
compared to \emph{bimonoidal categories}.  As is usual with type theories,
the coherence conditions (pentagonal and triangle identity) are not explicit,
but would correspond to certain identity types being trivial.

@@

The language {{Pi}} models isomorphisms of values rather well.  However,
purposefully, it has no distinguished notion of \emph{input} or
\emph{output}; more precisely, because of strict preservation of information,
these two concepts coincide.  This is the root of reversibility. But if we
want to model functions of any flavor as values, we need to differentiate
between these notions. We also clearly need to relax the strict separation
between values and isomorphisms as the notion of ``higher-order'' requires
that isomorphisms be represented as values. Our idea, inspired by
computational dualities~\cite{Filinski:1989:DCI:648332.755574,Curien:2000},
polarity~\cite{Girard87tcs,10.1109/LICS.2010.23}, and the categorical notion
of \emph{compact
  closure}~\cite{Selinger:2007:DCC:1229185.1229207,Abramsky:2004:CSQ:1018438.1021878},
is to introduce a \emph{formal dual} for our types.  We thus consider a type
in a negative position to possess negative information, as it is really a
request for information.  Since information is logarithmic, this means that a
request for information should behave like a \emph{fractional type}.

@@

The extension of the sets of types and values from {{Pi}} to {{PiTF}} is
simple:
%subcode{bnf} include main
% value types, b ::= ... | 1/b
% values, v ::= ... | 1/v
For a given type {{b}}, the values of type {{1/b}} are of the form {{1/v}}
where {{v : b}}. Note that {{1/v}} is purely formal, as is {{1/b}}. We have
however some informal intuition about the \emph{meaning} of such fractional
types and values. (See Sec.~\ref{sec:cat} for more discussion about this
point.) In more detail, given a symmetric monoidal category modeling a
certain language (like {{Pi}}), one can obtain a higher-order extension of
the language by extending the category to a \emph{compact closed} one, i.e.,
a category in which morphisms are representable as objects. The key to this
extension is to add dual objects (in our case the fractionals) satisfying the
following isomorphism witnessed by two new type indexed combinators 
{{eta*_b}} and {{eps*_b}}:
\begin{multicols}{3}
  \dgm{eta_times2}

@@

From the categorical perspective, versions of compact closed categories were
introduced by Abramsky and Coecke~\cite{Abramsky:2004:CSQ:1018438.1021878}
and by Selinger~\cite{Selinger:2007:DCC:1229185.1229207} as the
generalization of monoidal categories to model quantum computation.  The
first and most intuitive example of such categories is the category of finite
sets and relations. In that case, the dual operation (i.e., the fractional
type) collapses by defining {{1/b}} to be {{b}}. In spite of this degenerate
interpretation of fractionals, this is still an interesting category as it
provides a model for a reversible higher-order language. Indeed, any relation
is reversible via the standard \emph{converse} operation on a relation.
Furthermore, as any relation between {{b1}} and {{b2}} can be
represented as a subset of {{b1 * b2}}, i.e., as a value. A more interesting
example is the category of finite dimensional vector spaces and linear maps
(or even the category of finite dimensional Hilbert spaces). In such
categories the dual operation takes a vector space (consisting of ``column
vectors'') to the dual vector space (consisting of ``row vectors''). While
these are isomorphic, but they are nevertheless quite different,
being $(1,0)$ and $(0,1)$ tensors (respectively).  In other words, the
category provides a more refined model in which isomorphic
negative-information values and positive-information are not identified.

@@

\begin{definition}[Denotation of Value Types {{ [[ b ]] }}]
\label{chx:def:denot}
Each type denotes a finite set of values as follows:
%subcode{opsem} include main
%! <- = \leftarrow
%! union = \cup
% [[1]] '= {[ () ]}
% [[b1 + b2]] '= {[ left v ~|~ v <- [[b1]] ]} union {[ right v ~|~ v <- [[b2]] ]}
% [[b1 * b2]] '= {[ (v1, v2) ~|~ v1 <- [[b1]], v2 <- [[b2]] ]}
% [[1/b]] '= {[ 1/v ~|~ v <- [[b]] ]}
\end{definition}

@@

\begin{definition}[Relational semantics]
\label{def:relational-PiTF}
Each combinator {{c : b1 <-> b2}} in {{PiTF}} denotes a relation
as specified below. 
%subcode{proof} include main
%@ ~
%@@ (left v) ~~swap+~~ (right v)
%
%@ ~
%@@ (right v) ~~swap+~~ (left v)
%----
%@ ~
%@@ (left v) ~~assocl+~~ (left (left v))
%
%@ ~
%@@ (right (left v)) ~~assocl+~~ (left (right v))
%----
%@ ~
%@@ (right (right v)) ~~assocl+~~ (right v)
%
%@ ~
%@@ (left (left v)) ~~assocr+~~ (left v)
%----
%@ ~
%@@ (left (right v)) ~~assocr+~~ (right (left v))
%
%@ ~
%@@ (right v) ~~assocr+~~ (right (right v))
%----
%@ ~
%@@ ((), v) ~~unite~~ v
%
%@ ~
%@@ v ~~uniti~~ ((), v)
%----
%@ ~
%@@ (v1, v2) ~~swap*~~ (v2, v1)
%----
%@ ~
%@@ (v1, (v2, v3)) ~~assocl*~~ ((v1, v2), v3)
%
%@ ~
%@@ ((v1, v2), v3) ~~assocr*~~ (v1, (v2, v3))
%----
%@ ~
%@@ (left v1, v3) ~~dist~~ (left (v1, v3))
%
%@ ~
%@@ (right v2, v3) ~~dist~~ (right (v2, v3))
%----
%@ ~
%@@ (left (v1, v3)) ~~factor~~ (left v1, v3)
%
%@ ~
%@@ (right (v2, v3)) ~~factor~~ (right v2, v3)
%----
%@ ~
%@@ v ~~id~~ v
%
%@ v' ~~c~~ v
%@@ v ~~(sym c)~~ v'
%
%@ v1 ~~c1~~ v2
%@ v2 ~~c2~~ v3
%@@ v1 ~~(c1(;)c2)~~ v3
%----
%@ v ~~c1~~ v'
%@@ (left v) ~~(c1 (+) c2)~~ (left v')
%
%@ v ~~c2~~ v'
%@@ (right v) ~~(c1 (+) c2)~~ (right v')
%----
%@ v1 ~~c1~~ v1'
%@ v2 ~~c2~~ v2'
%@@ (v1, v2) ~~(c1 (*) c2)~~ (v1', v2')
%----
%@ ~
%@@ () ~~eta*_b~~ (1/v,v)
%
%@ ~
%@@ (1/v,v) ~~eps*_b~~ ()
\end{definition}

The operational semantics define a relation rather than a function, but the
goal is for them to still be reversible; in particular, we wish for the
relation to mode check so that for any combinator {{c}} we can treat either
its left or its right side as an input; this allows us to define {{sym}} as
just relational converse, which amounts to
an argument swap. We can then give an operational interpretation to the
semantic relation by defining two functions: a forward evaluator and a
backwards evaluator that exercise the relation in opposite directions,
but which now can return a \emph{set} of results. It is
straightforward to check that if the forward evaluation of {{c v1}} produces
{{v2}} as a possible answer then the backwards evaluation of {{c v2}}
produces {{v1}} as a possible answer.  One can check that all relations
for {{Pi}} are total functional relations whose converse are also total 
functional relations -- aka isomorphisms.  However {{eta*_b}} relates
{{()}} to multiple values (one for each value of {{b}}), and is thus
not functional; {{eps*_b}} is exactly its relational converse, as expected.

In the interest of brevity, we have treated the type annotations in the base
isomorphisms as implicit; for example, the rule for {{assocr*}} should really
read:

%subcode{proof} include main
%@ v1 in b1
%@ v2 in b2
%@ v3 in b3
%@@ ((v1, v2), v3) ~~assocr*_{b1, b2, b3}~~ (v1, (v2, v3))

@@

In the following, we use {{bool}} as an abbreviation of {{1+1}} 
with {{true}} as an abbreviation for {{left ()}} and {{false}} as an 
abbreviation for {{right ()}}. In addition, for the sake of readability, the
circuit diagrams in this section will elide trivial unit wires, such as in
{{eta}} and {{epsilon}}.

@@

As motivated, with fractional types, we can define a value that denotes a
first-class function or relation. Indeed a value of type {{1/b1 * b2}} is a
pair of a \emph{constraint} that can only be satisfied by some {{v1 : b1}}
and a value {{v2 : b2}}. In other words, it corresponds to a function or
relation which when ``given'' a value {{v1 : b1}} ``releases'' the value 
{{v2 : b2}}. To emphasize this view, we introduce the abbreviation:

@@

\noindent which suggests a function-like behavior for the pair of a fractional value
and a regular value. What is remarkable is that we can almost trivially turn
any combinator {{c : b1 <-> b2}} into a (constant) value of 
type {{b1 :-* b2}} as shown below:

@@

  \dgm{delimc-times}
\end{multicols}
\noindent  Dually, as illustrated above, we also have a {{coname}} combinator
which when given an answer {{v2 : b2}}, and a request for the corresponding
input {{1/b1}} for {{c : b1 <-> b2}}, eliminates this as a un-needed 
computation.

As a simple example, consider {{swap+}} denoting boolean
negation.  It is straightforward to check that the combinator 
{{name~swap+ : 1 <-> (bool :-* bool)}} denotes the following relation:

%subcode{proof} include main
%@ ~
%@@ () ~~(name~swap+)~~ (1/false,true)
%
%@ ~
%@@ () ~~(name~swap+)~~ (1/true,false)

\noindent The relation {{name~swap+}} maps {{()}} to a value which can be used
to convert any {{false}} to {{true}} and vice-versa. We formalize this
``apply'' combinator below.

@@

We are now a small step from implementing various higher-order
combinators that manipulate functions or relations. In particular, we
can \emph{apply} and \emph{compose} values representing relations.
\emph{apply} is shown below:

@@

%subcode{opsem} include main
%! columnStyle = rcl
% curry &:& b1 * b2 :-* b3 <-> b1 :-* (b2 :-* b3)
% curry &=& (id * recip) (;) (id * swap*) (;) assocl*
\noindent That {{recip}} is the heart of currying seems quite remarkable.
%%%%%%%%%%%%
\subsection{Duality}

The final remarkable property of fractional types is that they
are involutive:

@@

Mathematically speaking, recursion and iteration can be expressed using
categorical trace
operators~\cite{joyal1996traced,Hasegawa:1997:RCS:645893.671607}.  In a
language like {{Pi}} there are two natural families of trace operators
that can be defined, an additive family (explored in detail in our previous
work~\cite{rc2011}) and a multiplicative family, which we will focus on now.

Given a computation {{c:b1 * b2 <-> b1 * b3}}, we
want to cancel the common type {{b1}}, to produce a new combinator
{{trace*~c:b2<->b3}}. For its evaluation, we are only
given a value {{v2 : b2}}. But to evaluate {{c}}, we must provide a pair
consisting of ``some'' value {{v1 : b1}} together with the given {{v2 : b2}}.
The needed value {{v1 : b1}} cannot be arbitrary: it must be such that it
is the value produced as the first component of the result.  In general, there
may be several such fixed-point values, or none.  A few examples should help
explain the subtleties.

\begin{example}
Consider the combinator {{c : bool * 1 <-> bool * 1}}, with {{c = id}}.

\noindent Using {{trace*}}, we construct the
combinator {{trace* c : 1 <-> 1}}. Applying this to 
{{() : 1}} requires us to find a value {{b : bool}} such that 
{{(b,()) ~~d~~ (b,())}}. Given that the type {{bool}} 
has two values, there are two values {{false}} and {{true}} that satisfy the 
constraint. In other words, {{() ~~(trace* c)~~() }}.
\end{example}

\begin{example}
\label{ch3:ex:annihilate}
Consider a small variation on the combinator {{c}} above:

{{c : bool * 1 <-> bool * 1}}

{{c = swap+ (*) id }}

\noindent which negates the boolean component of its incoming
pair. Using the multiplicative trace operator, we can construct the
combinator {{trace* c : 1 <-> 1}} as before. But now, applying this
combinator to {{() : 1}} requires us to find a value {{b : bool}} such
that {{(b,()) ~~swap+~~ (b,())}} which is impossible since {{swap+}} 
has no fixed points.  Operationally, the evaluation of
such a combinator should produce no value (were it to even terminate).
\end{example}

@@

We can use the relational semantics of definition~\ref{def:relational-PiTF}
to calculate the result of applying
{{trace*_{bool} (swap+ (*) id)}} to {{false}}.  We use the usual conversion
of a relation from a subset of {{A * B}} to a function from {{A}} to the
powerset of {{B}} (and Kleisli composition in the powerset monad) for
the computation:
%subcode{opsem} include main
%! columnStyle = ll
% {[ ((),false) ]}  & (uniti)
% {[ ((1/false,false),false), ((1/true,true),false) ]} & (eta*_{bool} * id)
% {[ (1/false, (false,false)), (1/true,(true,false)) ]} & (assocr*)
% {[ (1/false, (true,false)), (1/true,(false,false)) ]} & (id (*) not )
% {[ ((1/false,true),false), ((1/true,false),false) ]} & (assocl*)
% emptyset & (eps*_{bool} * id)

\noindent  The computation with {{true}} gives the same result.  This
implies that although our semantics is formally reversible, it 
does not result in isomorphisms.
Nevertheless, the next subsection illustrates why this is still useful.

@@

\noindent We can define a combinator {{c_R}} whose denotation is {{R}},
by defining a combinator {{cInner : (a * bool) <-> (a * bool)}}
for some type {{a}} such that {{c_R = trace* cInner}}:

@@

%subcode{opsem} include main
%! columnStyle = rcl
% cInner &:& (bool * bool) * bool <-> (bool * bool) * bool
% cInner &=& sym (swap* (;) assocl* (;) (cnot (*) id) (;) assocr* (;) swap* (;)  toffoli (;)
%        &&  ((swap* (;) cnot (;) swap*) (*) id) (;) toffoli (;)
%        &&  (assocr* (;) swap* (;) toffoli (;) swap* (;) assocl*) (;) toffoli (;) (cnot (*) id))

>>>>

\paragraph*{Data.}
We view {{Pi}} as having two levels:  it has traditional values, given by:
%subcode{bnf} include main
% values, v ::= () | left v | right v | (v, v)

\noindent and these are classified by ordinary types:
%subcode{bnf} include main
% value types, b ::= 1 | b + b | b * b

\noindent 
Types include the unit type {{1}}, sum types {{b1+b2}}, and
products types {{b1*b2}}.  Values includes {{()}} which is the only value of
type {{1}}, {{left v}} and {{right v}} which inject~{{v}} into a sum type,
and {{(v1,v2)}} which builds a value of product type. But these values should
be regarded as largely ancillary: we do not treat them as first-class citizens,
and they only occur when observing the effect of an isomorphism.

\paragraph*{Isomorphisms.} The important terms of {{Pi}} are witnesses to
(value) type isomorphisms.  They are also typed, by the shape of the (value)
type isomorphism they witness {{b <-> b}}.  Specifically, they are witnesses to
the following isomorphisms:
%subcode{bnf} include main
%! columnStyle = r@{\hspace{-0.5pt}}r c l@{\hspace{-0.5pt}}l
%swap+ :&  b1 + b2 & <-> & b2 + b1 &: swap+
%assocl+ :&  b1 + (b2 + b3) & <-> & (b1 + b2) + b3 &: assocr+
%identl* :&  1 * b & <-> & b &: identr*
%swap* :&  b1 * b2 & <-> & b2 * b1 &: swap*
%assocl* :&  b1 * (b2 * b3) & <-> & (b1 * b2) * b3 &: assocr*
%dist :&~ (b1 + b2) * b3 & <-> & (b1 * b3) + (b2 * b3)~ &: factor

\noindent Each line of the above table introduces a pair of dual
constants\footnote{where {{swap*}} and {{swap+}} are self-dual.} that witness
the type isomorphism in the middle.  These are the base (non-reducible) terms
of the second, principal level of {{Pi}}. Note how the above has two
readings: first as a set of typing relations for a set of constants. Second,
if these axioms are seen as universally quantified, orientable statements,
they also induce transformations of the (traditional) values. The
(categorical) intuition here is that these axioms have computational content
because they witness isomorphisms rather than merely stating an extensional
equality.

@@

It is important to note that ``values'' and ``isomorphisms'' are completely
separate syntactic categories which do not intermix. The semantics of the
language come when these are made to interact at the ``top level'' via
\emph{application}:
%subcode{bnf} include main
% top level term, l ::= c v

@@

The language presented above, at the type level, models a commutative ringoid
where the multiplicative structure forms a commutative monoid, but the
additive structure is just a commutative semigroup.  Note that the version of
{{Pi}} that include the empty type with its usual laws exactly captures, at
the type level, the notion of a \emph{semiring} (occasionally called a
\emph{rig}) where we replace equality by isomorphism.  Semantically, {{Pi}}
models a \emph{bimonoidal category} whose simplest example is the category of
finite sets and bijections. In that interpretation, each value type denotes a
finite set of a size calculated by viewing the types as natural numbers and
each combinator {{c : b1 <-> b2}} denotes a bijection between the sets
denoted by~{{b1}} and~{{b2}}. (We discuss the operational semantics in
conjuction with our extension in the next section.)

% Operationally, the
% semantics of {{Pi}} is given using two mutually recursive interpreters: one
% going forward and one going backwards. The use of {{sym}} switches control
% from one evaluator to the other. We will present the operational semantics in
% the next section along with the extension with fractional types and
% values. For now, we state without proof that the evaluation of well-typed
% combinators always terminates and that {{Pi}} is logically reversible, i.e.,
% that for all combinators {{c : b1 <-> b2}} and values {{v1 : b1}} and {{v2 :
%     b2}} we have the forward evaluation of {{c v1}} produces {{v2}} iff the
% backwards evaluation of {{c v2}} produces {{v1}}.

@@

%% As is usual with type theories,
%% the coherence conditions (pentagonal and triangle identity) are not explicit,
%% but would correspond to certain identity types being trivial.

@@

The language {{Pi}} models isomorphisms of values rather well, and as we
established before~\cite{James:2012:IE:2103656.2103667,rc2011,rc2012}, it is
logically reversible and each computation it expresses preserves the
information-theoretic entropy. However, purposefully, it has no distinguished
notion of \emph{input} or \emph{output}; more precisely, because of strict
preservation of information, these two concepts coincide.  This is the root
of reversibility. But if we want to model functions of any flavor as values,
we need to differentiate between these notions. Our idea, inspired by
computational dualities~\cite{Filinski:1989:DCI:648332.755574,Curien:2000},
polarity~\cite{Girard87tcs,10.1109/LICS.2010.23}, and the categorical notion
of \emph{compact
  closure}~\cite{Selinger:2007:DCC:1229185.1229207,Abramsky:2004:CSQ:1018438.1021878},
is to introduce a \emph{formal dual} for our types.  We thus consider a type
in a negative position to possess negative information, as it is really a
request for information.  Since information is logarithmic, this means that a
request for information should behave like a \emph{fractional type}.

@@

The extension of the sets of types and values from {{Pi}} to {{PiTF}} is
simple:
%subcode{bnf} include main
% value types, b ::= ... | 1/b
% values, v ::= ... | 1/v

\noindent For a given type {{b}}, the values of type {{1/b}} are of the form
{{1/v}} where {{v : b}}. Note that {{1/v}} is purely formal, as is
{{1/b}}.

Semantically, we are extending the symmetric monoidal category modeling
{{Pi}} to a \emph{compact closed} one, i.e., a category in which morphisms
are representable as objects. In such a setting, the new dual objects (i.e.,
the fractionals) must satisfy the following isomorphism witnessed by two new
type indexed combinators {{eta*_b}} and {{eps*_b}}:
\begin{multicols}{3}
  \dgm{eta_times2}

@@

Historically, versions of compact closed categories were introduced by
Abramsky and Coecke~\cite{Abramsky:2004:CSQ:1018438.1021878} and by
Selinger~\cite{Selinger:2007:DCC:1229185.1229207} as the generalization of
monoidal categories to model quantum computation.  The first and most
intuitive example of such categories is the category of finite sets and
relations. In that case, the dual operation (i.e., the fractional type)
collapses by defining {{1/b}} to be {{b}}. In spite of this degenerate
interpretation of fractionals, this is still an interesting category as it
provides a model for a reversible higher-order language. Indeed, any relation
is reversible via the standard \emph{converse} operation on a relation.
Furthermore, as any relation between {{b1}} and {{b2}} can be represented as
a subset of {{b1 * b2}}, i.e., as a value.

@@

\begin{definition}[Denotation of Value Types]
\label{chx:def:denot}
Each type denotes a finite set of values as follows:
%subcode{opsem} include main
%! <- = \leftarrow
%! union = \cup
% [[1]] '= {[ () ]}
% [[b1 + b2]] '= {[ left v ~|~ v <- [[b1]] ]} union {[ right v ~|~ v <- [[b2]] ]}
% [[b1 * b2]] '= {[ (v1, v2) ~|~ v1 <- [[b1]], v2 <- [[b2]] ]}
% [[1/b]] '= {[ 1/v ~|~ v <- [[b]] ]}
\end{definition}

@@

\begin{definition}[Relational semantics]
\label{def:relational-PiTF}
Each combinator {{c : b1 <-> b2}} in {{PiTF}} denotes a relation
as specified below.~\footnote{
In the interest of brevity, we have treated the type annotations in the base
isomorphisms as implicit; for example, the rule for {{assocr*}} should really
read:

%subcode{proof} include main
%@ v1 in b1
%@ v2 in b2
%@ v3 in b3
%@@ ((v1, v2), v3) ~~assocr*_{b1, b2, b3}~~ (v1, (v2, v3))
}
%subcode{proof} include main
%@ ~
%@@ (left v) ~~swap+~~ (right v)
%
%@ ~
%@@ (right v) ~~swap+~~ (left v)
%----
%@ ~
%@@ (left v) ~~assocl+~~ (left (left v))
%
%@ ~
%@@ (right (left v)) ~~assocl+~~ (left (right v))
%----
%@ ~
%@@ (right (right v)) ~~assocl+~~ (right v)
%
%@ ~
%@@ (left (left v)) ~~assocr+~~ (left v)
%----
%@ ~
%@@ (left (right v)) ~~assocr+~~ (right (left v))
%
%@ ~
%@@ (right v) ~~assocr+~~ (right (right v))
%----
%@ ~
%@@ ((), v) ~~unite~~ v
%
%@ ~
%@@ v ~~uniti~~ ((), v)
%----
%@ ~
%@@ (v1, v2) ~~swap*~~ (v2, v1)
%----
%@ ~
%@@ (v1, (v2, v3)) ~~assocl*~~ ((v1, v2), v3)
%
%@ ~
%@@ ((v1, v2), v3) ~~assocr*~~ (v1, (v2, v3))
%----
%@ ~
%@@ (left v1, v3) ~~dist~~ (left (v1, v3))
%
%@ ~
%@@ (right v2, v3) ~~dist~~ (right (v2, v3))
%----
%@ ~
%@@ (left (v1, v3)) ~~factor~~ (left v1, v3)
%
%@ ~
%@@ (right (v2, v3)) ~~factor~~ (right v2, v3)
%----
%@ ~
%@@ v ~~id~~ v
%
%@ v' ~~c~~ v
%@@ v ~~(sym c)~~ v'
%
%@ v1 ~~c1~~ v2
%@ v2 ~~c2~~ v3
%@@ v1 ~~(c1(;)c2)~~ v3
%----
%@ v ~~c1~~ v'
%@@ (left v) ~~(c1 (+) c2)~~ (left v')
%
%@ v ~~c2~~ v'
%@@ (right v) ~~(c1 (+) c2)~~ (right v')
%----
%@ v1 ~~c1~~ v1'
%@ v2 ~~c2~~ v2'
%@@ (v1, v2) ~~(c1 (*) c2)~~ (v1', v2')
%----
%@ ~
%@@ () ~~eta*_b~~ (1/v,v)
%
%@ ~
%@@ (1/v,v) ~~eps*_b~~ ()
\end{definition}

The semantics above defines a relation rather than a function. It is still
reversible in the sense that it ``mode-checks''. In other words, for any
combinator {{c}} we can treat either its left or its right side as an input;
this allows us to define {{sym}} as just relational converse, which amounts
to an argument swap. We can then give an operational interpretation to the
semantic relation by defining two functions: a forward evaluator and a
backwards evaluator that exercise the relation in opposite directions, but
which now can return a \emph{set} of results. This operational view exploits
the usual conversion of a relation from a subset of {{A * B}} to a function
from {{A}} to the powerset of {{B}} whose composition is given by the Kleisli
composition in the powerset monad. It is straightforward to check that if the
forward evaluation of {{c v1}} produces {{v2}} as a possible answer then the
backwards evaluation of {{c v2}} produces {{v1}} as a possible answer.  One
can check that all relations for {{Pi}} are total functional relations whose
converse are also total functional relations -- aka isomorphisms.  However
{{eta*_b}} relates {{()}} to multiple values (one for each value of {{b}}),
and is thus not functional; {{eps*_b}} is exactly its relational converse, as
expected.

@@

Having introduced the syntax of our extended language, its semantics in the
category of sets and relations, and its operational semantics using forward
and backwards interpreters, we now illustrate its expressiveness as a
higher-order programming language. In the following presentation consisting
of numerous programming examples, we use {{bool}} as an abbreviation of
{{1+1}} with {{true}} as an abbreviation for {{left ()}} and {{false}} as an
abbreviation for {{right ()}}. In addition, instead of presenting the code
using the syntax of {{Pi}}, we use circuit diagrams that are hopefully more
intuitive. Each diagrams represents a combinator whose evaluation consists of
propagating values along the wires.  For the sake of readability, we omit
obvious re-shuffling circuitry and elide trivial unit wires which carry no
information.\footnote{The full code is available upon request. We plan on
  releasing it as soon as possible.}

@@

The most basic additional expressiveness of {{PiTF}} over {{Pi}} is the
ability to express relations as values. Indeed a value of type {{1/b1 * b2}}
is a pair of a \emph{constraint} that can only be satisfied by some 
{{v1 : b1}} and a value {{v2 : b2}}. In other words, it corresponds to a
function or relation which when ``given'' a value {{v1 : b1}} ``releases''
the value {{v2 : b2}}. To emphasize this view, we introduce the abbreviation:

@@

\noindent which suggests a function-like behavior for the pair of a fractional value
and a regular value. What is remarkable is that we can almost trivially turn
any combinator {{c : b1 <-> b2}} into a (constant) value of 
type {{b1 :-* b2}} as shown below on the left:

@@

  \dgm{delimc-times}
\end{multicols}
\noindent Dually, as illustrated above on the right, we also have a
{{coname}} combinator which when given an answer {{v2 : b2}}, and a request
for the corresponding input {{1/b1}} for {{c : b1 <-> b2}}, eliminates this
as a un-needed computation.

More generally any combinator manipulating values of type {{b}} can be turned
into a combinator manipulating values of type {{1/b}} and vice-versa. This is
due to the fact that fractionals types satisfy a self-dual involution
relating {{b}} and {{1/(1/b)}}:

% As a simple example, consider {{swap+}} denoting boolean
% negation.  It is straightforward to check that the combinator 
% {{name~swap+ : 1 <-> (bool :-* bool)}} denotes the following relation:

% %subcode{proof} include main
% %@ ~
% %@@ () ~~(name~swap+)~~ (1/false,true)
% %
% %@ ~
% %@@ () ~~(name~swap+)~~ (1/true,false)

% \noindent The relation {{name~swap+}} maps {{()}} to a value which can be used
% to convert any {{false}} to {{true}} and vice-versa. We formalize this
% ``apply'' combinator below.

@@

We are now a small step from implementing various higher-order combinators
that manipulate functions or relations. In particular, we can \emph{apply},
\emph{compose}, \emph{curry}, and \emph{uncurry} values representing
relations.  We first show the realization of \emph{apply}:

@@

%subcode{opsem} include main
%! columnStyle = rcl
% curry &:& b1 * b2 :-* b3 <-> b1 :-* (b2 :-* b3)
% curry &=& (id * recip) (;) (id * swap*) (;) assocl*

\noindent That {{recip}} is the heart of currying seems quite remarkable.

@@

Mathematically speaking, recursion and iteration can be expressed using
categorical trace
operators~\cite{joyal1996traced,Hasegawa:1997:RCS:645893.671607}.  In a
language like {{Pi}} there are two natural families of trace operators that
can be defined, an additive family (explored in detail in our previous
work~\cite{rc2011}) and a multiplicative family, which is expressible using
fractionals.

The idea of the multiplicative trace operator is as follows. We are given a
computation {{c : b1 * b2 <-> b1 * b3}}, and we build a ``looping'' version
which feeds the output value of type {{b1}} back as an input. Effectively,
this construction cancels the common type {{b1}}, to produce a new combinator
{{trace*~c:b2<->b3}}.

@@

As an example, we can use the operational semantics (outlined in the previous
section) to calculate the result of applying {{trace*_{bool} (swap+ (*) id)}}
to {{false}}.
%subcode{opsem} include main
%! columnStyle = ll
% {[ ((),false) ]}  & (uniti)
% {[ ((1/false,false),false), ((1/true,true),false) ]} & (eta*_{bool} * id)
% {[ (1/false, (false,false)), (1/true,(true,false)) ]} & (assocr*)
% {[ (1/false, (true,false)), (1/true,(false,false)) ]} & (id (*) not )
% {[ ((1/false,true),false), ((1/true,false),false) ]} & (assocl*)
% emptyset & (eps*_{bool} * id)

\noindent The computation with {{true}} gives the same result. This confirms
that although our semantics is formally reversible, it does not result in
isomorphisms.

More abstractly, the evaluation of {{trace*~c : b2 <-> b3}} must ``guess'' a
value of type {{b1}} to be provided to to the inner combinator 
{{c : b1 * b2 <-> b1 * b3}}. 
This value {{v1 : b1}} cannot be arbitrary: it must be such that it is the value
produced as the first component of the result.  In general, there may be
several such fixed-point values, or none. Indeed with a little ingenuity 
(see the next section), one can 
express any desired relation by devising a circuit which keeps the desired pairs as 
valid fixed-points.

% \begin{example}
% Consider the combinator {{c : bool * 1 <-> bool * 1}}, with {{c = id}}.

% \noindent Using {{trace*}}, we construct the
% combinator {{trace* c : 1 <-> 1}}. Applying this to 
% {{() : 1}} requires us to find a value {{b : bool}} such that 
% {{(b,()) ~~d~~ (b,())}}. Given that the type {{bool}} 
% has two values, there are two values {{false}} and {{true}} that satisfy the 
% constraint. In other words, {{() ~~(trace* c)~~() }}.
% \end{example}

% \begin{example}
% \label{ch3:ex:annihilate}
% Consider a small variation on the combinator {{c}} above:

% {{c : bool * 1 <-> bool * 1}}

% {{c = swap+ (*) id }}

% \noindent which negates the boolean component of its incoming
% pair. Using the multiplicative trace operator, we can construct the
% combinator {{trace* c : 1 <-> 1}} as before. But now, applying this
% combinator to {{() : 1}} requires us to find a value {{b : bool}} such
% that {{(b,()) ~~swap+~~ (b,())}} which is impossible since {{swap+}} 
% has no fixed points.  Operationally, the evaluation of
% such a combinator should produce no value (were it to even terminate).
% \end{example}

@@

\noindent We can define a combinator {{c_R}} whose denotation is {{R}}, by
defining a combinator {{cInner : (a * bool) <-> (a * bool)}} for some type
{{a}} such that {{c_R = trace* cInner}}. The basic requirement of {{cInner}}
is that for each desired pair {{(v1,v2)}} in {{R}}, it maps {{(a0,v1)}} to
{{(a0,v2)}} for some value {{a0}} and that for each pair {{(v1,v2)}} that is
\emph{not} in {{R}}, it maps {{(a1,v1)}} to {{(a2,v2)}} for \emph{different}
values {{a1}} and {{a2}}.

@@

% %subcode{opsem} include main
% %! columnStyle = rcl
% % cInner &:& (bool * bool) * bool <-> (bool * bool) * bool
% % cInner &=& sym (swap* (;) assocl* (;) (cnot (*) id) (;) assocr* (;) swap* (;)  toffoli (;)
% %        &&  ((swap* (;) cnot (;) swap*) (*) id) (;) toffoli (;)
% %        &&  (assocr* (;) swap* (;) toffoli (;) swap* (;) assocl*) (;) toffoli (;) (cnot (*) id))

% \noindent where {{toffoli}} and {{cnot}} are the conventional reversible gates (their definition in {{Pi}} is in~\cite{rc2011}).
% %     
% % c_R &:& bool <-> bool
% % c_R &=& trace* cInner
